---
title: "Income and Wildfire Risk Analysis"
author: "Jiayuan Jiang"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: bootstrap
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Introduction

This report analyzes the relationship between household income and wildfire risk across California census block groups.
The goal is to understand whether income level is associated with fire hazard exposure, and how this relationship varies spatially.

We use a dataset that combines

- Median household income from the ***American Community Survey (ACS 2022)***

- Fire Hazard Severity Zones from ***CAL FIRE***

***

###Data Preparation
```{r}
pacman::p_load(
  sf,           # Spatial data processing
  ggplot2,      # Plotting
  dplyr,        # Data processing
  scales,       # Scales and formatting
  viridis,      # Color mapping
  ggspatial,    # Add compass and scale bar
  RColorBrewer, # Color schemes
  classInt,     # Classification intervals
  readr,
  scales,
  tmap
)
# Set working directory (adjust to your path)

setwd("/Users/avalanchy/Downloads/EAS548-648-Geovisiual/lab02")

# Read and clean data

data <- read_csv("census_fire_analysis/census_fire_risk_data.csv", show_col_types = FALSE) %>%
filter(!is.na(income_clean),!is.na(fire_risk_level),income_clean > 0,!is.na(income_quantile_label)) %>%
mutate(fire_risk_category = factor(fire_risk_level,levels = 0:3,labels = c("No Risk", "Low Risk", "Moderate Risk", "High Risk")),income_quantile_ordered =factor(income_quantile_label,levels = c("Lowest 20%", "Low 20%", "Middle 20%", "High 20%", "Highest 20%")),income_k = income_clean / 1000)

summary(data)
```

We clean the dataset by removing missing or invalid values and create two main variables

-`income_quantile_ordered`: categorical income quintiles

-`fire_risk_category`: four fire risk levels (No, Low, Moderate, High)

***

### Geovisualization: Fire Risk Map
```{r}
# Census Income Data and Fire Risk Visualization
pacman::p_load(
  sf,           # Spatial data processing
  ggplot2,      # Plotting
  dplyr,        # Data processing
  scales,       # Scales and formatting
  viridis,      # Color mapping
  ggspatial,    # Add compass and scale bar
  RColorBrewer, # Color schemes
  classInt      # Classification intervals
)

setwd("/Users/avalanchy/Downloads")

# Define main function
visualize_census_fire_risk <- function(
    census_file = "/Users/avalanchy/Downloads/FHSZ_SRA_LRA_Combined_-2127032115786874601.geojson",
    fire_risk_file = "census_with_fire_risk.geojson",
    income_field = "X19_INCOME_B19013_E001",
    output_file = "census_fire_risk_map.png",
    width = 16, 
    height = 12,
    dpi = 300
) {
  
  cat("Reading data files...\n")
  
  # Read census income data
  census_data <- st_read(census_file, quiet = TRUE)
  cat(sprintf("Census data: %d records\n", nrow(census_data)))
  
  # Read fire risk data
  fire_risk_data <- st_read(fire_risk_file, quiet = TRUE)
  cat(sprintf("Fire risk data: %d records\n", nrow(fire_risk_data)))
  
  # Check if income field exists
  if (!income_field %in% names(census_data)) {
    cat("Error: Income field does not exist\n")
    cat("Available fields:\n")
    print(names(census_data))
    return(NULL)
  }
  
  # Data preprocessing
  cat("Preprocessing data...\n")
  
  # Process census income data
  census_clean <- census_data %>%
    mutate(
      income = as.numeric(.data[[income_field]]),
      income_clean = ifelse(income < 0 | is.na(income), NA, income)
    ) %>%
    filter(!is.na(income_clean))
  
  cat(sprintf("Cleaned Census data: %d records\n", nrow(census_clean)))
  cat(sprintf("Income statistics: Min=$%s, Max=$%s, Median=$%s\n", 
              format(min(census_clean$income_clean), big.mark=","),
              format(max(census_clean$income_clean), big.mark=","),
              format(median(census_clean$income_clean), big.mark=",")))
  
  # Income data quantile classification
  quantile_breaks <- quantile(census_clean$income_clean, 
                              probs = c(0, 0.2, 0.4, 0.6, 0.8, 1.0), 
                              na.rm = TRUE)
  
  census_clean <- census_clean %>%
    mutate(
      income_quantile = cut(income_clean, 
                            breaks = quantile_breaks, 
                            include.lowest = TRUE,
                            labels = c("Lowest 20%", "Low 20%", "Middle 20%", "High 20%", "Highest 20%"))
    )
  
  # Process fire risk data
  fire_risk_clean <- fire_risk_data %>%
    filter(fire_risk_level > 0) %>%  # Only show areas with fire risk
    mutate(
      fire_risk_factor = factor(fire_risk_level, 
                                levels = 1:3,
                                labels = c("Low Risk", "Moderate Risk", "High Risk"))
    )
  
  cat(sprintf("Areas with fire risk: %d records\n", nrow(fire_risk_clean)))
  if (nrow(fire_risk_clean) > 0) {
    print(table(fire_risk_clean$fire_risk_factor))
  }
  
  # Calculate centroids for point display
  fire_risk_centroids <- fire_risk_clean %>%
    st_centroid() %>%
    mutate(
      point_size = case_when(
        fire_risk_level == 1 ~ 0.3,
        fire_risk_level == 2 ~ 1.3,
        fire_risk_level == 3 ~ 2.0
      )
    )
  
  # Ensure coordinate systems are consistent
  if (st_crs(census_clean) != st_crs(fire_risk_centroids)) {
    fire_risk_centroids <- st_transform(fire_risk_centroids, st_crs(census_clean))
  }
  
  # Create map
  cat("Creating map...\n")
  
  # Define color schemes - using different color families for income and fire risk
  income_colors <- brewer.pal(5, "Blues")  # Blue gradient for income
  fire_colors <- c("Low Risk" = "#FED976", "Moderate Risk" = "#FD8D3C", "High Risk" = "#E31A1C")  # Yellow to red for fire risk
  
  # Calculate map boundaries
  bbox <- st_bbox(census_clean)
  
  # Create ggplot map
  p <- ggplot() +
    # Bottom layer: Census income data (polygons)
    geom_sf(data = census_clean, 
            aes(fill = income_quantile),
            color = "white", 
            size = 0.05,
            alpha = 0.8) +
    
    # Top layer: Fire risk data (points)
    geom_sf(data = fire_risk_centroids,
            aes(color = fire_risk_factor, 
                size = fire_risk_factor),
            alpha = 0.5,
            stroke = 0.3) +
    
    # Income color mapping
    scale_fill_manual(
      name = "Household Income\nQuantiles",
      values = income_colors,
      na.value = "grey95"
    ) +
    
    # Fire risk color and size mapping
    scale_color_manual(
      name = "Fire Risk Level",
      values = fire_colors
    ) +
    
    scale_size_manual(
      name = "Fire Risk Level",
      values = c("Low Risk" = 0.8, "Moderate Risk" = 1.5, "High Risk" = 2.5)
    ) +
    
    # Add scale bar only
    ggspatial::annotation_scale(
      location = "bl",
      unit_category = "metric",
      text_family = "Arial",
      text_col = "grey30",
      bar_cols = c("grey60", "white"),
      line_width = 1,
      height = unit(0.3, "cm"),
      text_cex = 0.8
    ) +
    
    # Map styling
    theme_void() +
    theme(
      legend.position = "right",
      legend.box = "vertical",
      legend.spacing.y = unit(0.5, "cm"),
      legend.margin = margin(l = 20),
      legend.title = element_text(size = 14, face = "bold"),  # Larger legend font
      legend.text = element_text(size = 12),                  # Larger legend font
      legend.key.size = unit(0.8, "cm"),
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      plot.margin = margin(10, 10, 10, 10)
    ) +
    
    # Coordinate system
    coord_sf(
      xlim = c(bbox[1], bbox[3]),
      ylim = c(bbox[2], bbox[4]),
      expand = FALSE
    )
  
  # Save map
  cat("Saving map...\n")
  ggsave(
    filename = output_file,
    plot = p,
    width = width,
    height = height,
    dpi = dpi,
    units = "in",
    bg = "white"
  )
  
  cat(sprintf("Map saved to: %s\n", output_file))
  
  # Display map
  print(p)
  
  # Return statistics
  stats <- list(
    census_records = nrow(census_clean),
    fire_risk_records = nrow(fire_risk_clean),
    income_summary = summary(census_clean$income_clean),
    income_quantiles = quantile_breaks,
    fire_risk_distribution = table(fire_risk_clean$fire_risk_factor)
  )
  
  return(stats)
}

# Create detailed statistics report function
print_statistics <- function(stats) {
  cat("\n=== Data Statistics Report ===\n")
  cat(sprintf("Census areas processed: %d\n", stats$census_records))
  cat(sprintf("Areas with fire risk: %d\n", stats$fire_risk_records))
  
  cat("\nIncome statistics (USD):\n")
  income_summary <- stats$income_summary
  cat(sprintf("  Minimum: $%s\n", format(income_summary[1], big.mark=",")))
  cat(sprintf("  1st Quartile: $%s\n", format(income_summary[2], big.mark=",")))
  cat(sprintf("  Median: $%s\n", format(income_summary[3], big.mark=",")))
  cat(sprintf("  Mean: $%s\n", format(income_summary[4], big.mark=",")))
  cat(sprintf("  3rd Quartile: $%s\n", format(income_summary[5], big.mark=",")))
  cat(sprintf("  Maximum: $%s\n", format(income_summary[6], big.mark=",")))
  
  cat("\nIncome quantile breakpoints:\n")
  for (i in 1:length(stats$income_quantiles)) {
    cat(sprintf("  %d%%: $%s\n", (i-1)*20, format(stats$income_quantiles[i], big.mark=",")))
  }
  
  cat("\nFire risk level distribution:\n")
  fire_dist <- stats$fire_risk_distribution
  for (i in 1:length(fire_dist)) {
    cat(sprintf("  %s: %d areas\n", names(fire_dist)[i], fire_dist[i]))
  }
}

# Usage example
if (TRUE) {  # Set to TRUE to run example
  
  # Set file paths
  census_file <- "/Users/avalanchy/Downloads/MedianIncome.geojson"
  fire_risk_file <- "/Users/avalanchy/Downloads/EAS548-648-Geovisiual/lab02/census_with_fire_risk.geojson"
  income_field <- "ACS_2022_5YR_BG_06_CALIFORNIA -X19_INCOME_B19013_E001"  # Adjust based on actual field name
  
  census_data <- st_read("/Users/avalanchy/Downloads/MedianIncome.geojson", quiet = TRUE)
  
  # Execute visualization
  tryCatch({
    result_stats <- visualize_census_fire_risk(
      census_file = census_file,
      fire_risk_file = fire_risk_file,
      income_field = names(census_data)[[52]],
      output_file = "census_fire_risk_visualization.png",
      width = 18,  # Increased width for larger map
      height = 14, # Increased height for larger map
      dpi = 300
    )
    
    # Print statistics
    if (!is.null(result_stats)) {
      print_statistics(result_stats)
    }
    
  }, error = function(e) {
    cat("Error:", e$message, "\n")
    cat("Please check:\n")
    cat("1. File paths are correct\n")
    cat("2. Field names are correct\n")
    cat("3. Required R packages are installed\n")
  })
}

# Optional: Create simplified version for quick testing
create_simple_map <- function(census_file, fire_risk_file) {
  census_data <- st_read(census_file, quiet = TRUE)
  fire_risk_data <- st_read(fire_risk_file, quiet = TRUE)
  
  ggplot() +
    geom_sf(data = census_data, fill = "lightgray", color = "white", size = 0.1) +
    geom_sf(data = st_centroid(fire_risk_data %>% filter(fire_risk_level > 0)),
            aes(color = factor(fire_risk_level)), size = 0.5) +
    scale_color_manual(values = c("1" = "yellow", "2" = "orange", "3" = "red")) +
    theme_void()
}
```
This map shows the ***spatial pattern*** of fire risk across California.
Areas in darker red represent higher fire hazard zones.
Combining this with income data helps visualize whether wealthier regions overlap with higher-risk areas.

***

### Statistical Relationship: Income vs Fire Risk
```{r}
# Scatter plot with loess trend line

data_filtered <- data %>%
filter(fire_risk_level > 0)

kendall_result <- cor.test(data_filtered$income_clean, data_filtered$fire_risk_level, method = "kendall")
tau_value <- round(as.numeric(kendall_result$estimate), 4)

p <- ggplot(data_filtered, aes(x = income_k, y = fire_risk_level)) +
geom_jitter(aes(color = fire_risk_category), alpha = 0.4, size = 1.5) +
geom_smooth(method = "loess", se = TRUE, color = "#2c3e50", fill = "#3182BD", linewidth = 1.2, alpha = 0.3) +
scale_color_manual(
name = "Fire Risk Level",
values = c("Low Risk" = "#FED976", "Moderate Risk" = "#FD8D3C", "High Risk" = "#E31A1C")
) +
scale_x_log10(
name = "Median Household Income ($1,000s)",
breaks = c(10, 20, 50, 100, 200, 500),
labels = function(x) paste0("$", x, "k")
) +
scale_y_continuous(
name = "Fire Risk Level",
breaks = 1:3,
labels = c("Low", "Moderate", "High")
) +
theme_minimal() +
labs(
title = "Relationship Between Income and Fire Risk",
subtitle = paste0("Kendall's τ = ", tau_value, "  (p = ", signif(kendall_result$p.value, 3), ")")
)

p
```
We use Kendall’s Tau correlation test to quantify the monotonic relationship between income and fire risk.

- ***Kendall’s τ*** = `r tau_value`

- A small positive value suggests that higher income areas tend to have ***slightly higher*** fire risk exposure, likely due to development patterns in suburban–wildland interface zones.


***

### Distribution Across Income Groups

```{r}
plot_data <- data %>%
group_by(income_quantile_ordered, fire_risk_category) %>%
summarise(count = n(), .groups = "drop") %>%
group_by(income_quantile_ordered) %>%
mutate(
total = sum(count),
proportion = count / total,
percentage = proportion * 100
)

ggplot(plot_data, aes(x = income_quantile_ordered, y = percentage, fill = fire_risk_category)) +
geom_col(position = "stack", width = 0.7) +
geom_text(
aes(label = ifelse(percentage > 5, paste0(round(percentage, 1), "%"), "")),
position = position_stack(vjust = 0.5),
color = "white",
fontface = "bold"
) +
scale_fill_manual(
name = "Fire Risk Level",
values = c(
"No Risk" = "#2E86AB",
"Low Risk" = "#A23B72",
"Moderate Risk" = "#F18F01",
"High Risk" = "#C73E1D"
)
) +
labs(
title = "Proportion of Fire Risk Levels Across Income Quantiles",
x = "Household Income Quantile",
y = "Percentage of Census Block Groups (%)"
) +
theme_minimal()

```
This stacked bar chart shows how fire risk levels distribute across income groups.
- We can observe that ***high-income block groups*** have a ***slightly higher*** percentage in the "High Risk" category, while lower-income groups are mostly in low-risk zones.
 ***
### Discussion
The analysis indicates a mild positive relationship between income and wildfire risk

- Wealthier households are often located in suburban areas near wildland–urban interfaces (WUI), where fire hazards are greater.

- Lower-income communities, more concentrated in urban centers, face lower wildfire exposure.

This aligns with findings from previous studies that ***environmental risk is not always inversely related to income***; some hazards disproportionately affect affluent regions due to land-use choices.

***

### Conclusion
Conclusion

- Dataset: ACS 2022 + CAL FIRE FHSZ

- Methods: Kendall’s Tau correlation and proportion visualization

- Main finding: Slight positive association between household income and fire hazard level

- Visualization tools: ggplot2, tmap

This project demonstrates how combining ***statistical analysis*** and ***geospatial visualization*** can uncover nuanced patterns between ***socioeconomic status*** and ***environmental risk***.
